A comparison of classification algorithms applied to the Adult Database

### 2009-12-14 report/final.tex
ordering of figures.
minor report updates.

There are 3620 records with missing values. The attributes that contain missing values are: \textbf{Workclass}, \textbf{Occupation} and \textbf{Native-country}.

@@

The table \ref{tab:summary} summarizes the distribution of values on the studied
dataset. From a first observation, one can see that the amount of missing values
is low. On the other side, in some of the attributes, some classes have a
much higher frequency than others, such as the 'united-states' native country, 
the 'private' workclass and the 'white' race. On the hours of work per week
attribute, there is a great amount of outliers, with some unrealistic
values, namely with amounts close to the 100 hours or less than 10 in 
individuals with a defined occupation, as can be seen on figure 
\ref{fig:box_hours}. Variables such has the capital gain and capital loss have
few values above 0, being quite specific to certain classes of individuals. The
sex is also unbalanced in the sample data, as the amount of men is more than
twice the amount of women.

@@

CN2 is an algorithm for inducing 
propositional classification rules. CN2 consists of two main procedures: the 
search procedure that performs beam search in order to find a single rule and 
the control procedure that repeatedly executes the search. 
The search procedure performs beam search using classification accuracy 
of the rule as a heuristic function.
Two different control procedures are used in CN2: one for inducing an ordered 
list of rules and the other for the unordered case.

@@

\begin{table}
  \begin{center}
  \begin{tabular}{ | l | c | c | c | c | c |}
    \hline
    \textbf{Type} & \textbf{Criterion} & \textbf{Train error rate} & \textbf{Test error rate} & \textbf{Time} \\ \hline
    ID3 & Gain Ratio & 18.38 & 18.61 & 0m04s \\ \hline
    ID3 & Information Gain & 18.48 & 19.05 & 0m03s \\ \hline
    ID3 & Gini Index & 18.94 & 19.15 & 0m04s \\ \hline
    ID3 & Accuracy & 20.41 & 20.24 & 0m04s \\ \hline
    ID3 Numerical & Gain Ratio & 14.93 & 15.68 & 0m14s \\ \hline
    ID3 Numerical & Information Gain & 15.68 & 16.42 & 0m06s \\ \hline
    ID3 Numerical & Gini Index & 16.10 & 16.96 & 0m07s \\ \hline
    ID3 Numerical & Accuracy & 17.99 & 17.92 & 0m12s \\ \hline
  \end{tabular}
  \caption{Results for ID3.}
  \label{tbl:results_id3}
  \end{center}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{ | l | l | l | l | l |}
    \hline
    \textbf{Algorithm} & \textbf{Error estimate} & \textbf{Train error rate} & \textbf{Test error rate} & \textbf{Time} \\ \hline
    Unordered & Laplacian & 13.3 & 15.2 & 3m29s \\ \hline
    Unordered & Naive & - & - & - \\ \hline
    Ordered & Laplacian & 0 & 17.6 & 1m45s \\ \hline
    Ordered & Naive & 0 & 19.9 & 12m21s \\ \hline
    Ordered & Information Gain & 18.9 & 18.9 & 0m5s \\ \hline
    Ordered & Modified Entropy & 18.9 & 19.1 & 0m20s \\ \hline
\end{tabular}
\caption{Results for CN2.}
\label{tbl:results_cn2}
\end{center}
\end{table}

@@

The error rate produced by Aleph when evaluating the test data was 20.45\%.
18881882 clauses were built, taking around 5 hours for the whole training
data. Despite the number of clauses, most of them are irrelevant, directly
classifying examples when no more general rule is available. Listing 
\ref{ilp_clauses} lists the most relevant clauses produced.

@@

\begin{table}
    \begin{center}
    \begin{tabular}{ | l | l | }
    \hline
    \textbf{Algorithm} & \textbf{Error rate} \\ \hline
    CN2 & 15.20 \\ \hline
    ID3 & 15.68 \\ \hline
    ADTree & 18.49 \\ \hline
    C4.5 & 18.84 \\ \hline
    Bayesian Networks (Simulated Annealing) & 19.15 \\ \hline
    Bayesian Networks (Hill Climbing) & 20.03 \\ \hline
    ILP (Aleph) & 20.45 \\ \hline
    Random Forest & 21.01 \\ \hline
    Bayesian Networks (Genetic Search) & 21.26 \\ \hline
    CART & 21.34 \\ \hline
    \end{tabular}
    \caption{Error rates for the different algorithms used in the classification process}
    \label{tbl:overall_results}
    \end{center}
\end{table}

@@

\caption{Summary of the value distribution in the database}

@@

>>>>

For the training set, there are 3620 records with missing values. The attributes
that contain missing values are: \textbf{Workclass}, \textbf{Occupation} and 
\textbf{Native-country}.

@@

The table \ref{tab:summary} summarizes the distribution of values in the training data on the studied
dataset, while the table \ref{tab:summarytest} summarizes the same distribution but
for the test set. A quick observation shows that the distribution of values is 
similar in both the training and the test set. Taking that into account, some
analysis was conducted on the training data, expecting some interesting 
properties to arise. From a first observation, one can see that the amount of missing values
is low. On the other side, in some of the attributes, some classes have a
much higher frequency than others, such as the 'united-states' native country, 
the 'private' workclass and the 'white' race. On the hours of work per week
attribute, there is a great amount of outliers, with some unrealistic
values, namely with amounts close to the 100 hours or less than 10 in 
individuals with a defined occupation, as can be seen on figure 
\ref{fig:box_hours}. Variables such has the capital gain and capital loss have
few values above 0, being quite specific to certain classes of individuals. The
sex is also unbalanced in the sample data, as the amount of men is more than
twice the amount of women.

@@

CN2 is an algorithm for inducing 
propositional classification rules. CN2 consists of two main procedures: the 
search procedure that performs beam search in order to find a single rule and 
the control procedure that repeatedly executes the search. 
The search procedure performs beam search using classification accuracy 
of the rule as a heuristic function.
Two different control procedures are used in CN2: one for inducing an ordered 
list of rules and the other for an unordered one.

@@

\begin{table}[ht]
  \begin{center}
  \begin{tabular}{ | l | c | c | c | c | c |}
    \hline
    \textbf{Type} & \textbf{Criterion} & \textbf{Train error rate} & \textbf{Test error rate} & \textbf{Time} \\ \hline
    ID3 & Gain Ratio & 18.38 & 18.61 & 0m04s \\ \hline
    ID3 & Information Gain & 18.48 & 19.05 & 0m03s \\ \hline
    ID3 & Gini Index & 18.94 & 19.15 & 0m04s \\ \hline
    ID3 & Accuracy & 20.41 & 20.24 & 0m04s \\ \hline
    ID3 Numerical & Gain Ratio & 14.93 & 15.68 & 0m14s \\ \hline
    ID3 Numerical & Information Gain & 15.68 & 16.42 & 0m06s \\ \hline
    ID3 Numerical & Gini Index & 16.10 & 16.96 & 0m07s \\ \hline
    ID3 Numerical & Accuracy & 17.99 & 17.92 & 0m12s \\ \hline
  \end{tabular}
  \caption{Results for ID3.}
  \label{tbl:results_id3}
  \end{center}
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{ | l | l | l | l | l |}
    \hline
    \textbf{Algorithm} & \textbf{Error estimate} & \textbf{Train error rate} & \textbf{Test error rate} & \textbf{Time} \\ \hline
    Unordered & Laplacian & 13.3 & 15.2 & 3m29s \\ \hline
    Unordered & Naive & - & - & - \\ \hline
    Ordered & Laplacian & 0 & 17.6 & 1m45s \\ \hline
    Ordered & Naive & 0 & 19.9 & 12m21s \\ \hline
    Ordered & Information Gain & 18.9 & 18.9 & 0m5s \\ \hline
    Ordered & Modified Entropy & 18.9 & 19.1 & 0m20s \\ \hline
\end{tabular}
\caption{Results for CN2.}
\label{tbl:results_cn2}
\end{center}
\end{table}

@@

The error rate produced by Aleph when evaluating the test data was 20.45\%.
Around 7000 clauses were built, taking around 5 hours for the whole training
data. Despite the number of clauses, most of them are irrelevant, directly
classifying examples when no more general rule is available. Listing 
\ref{ilp_clauses} lists the most relevant clauses produced.

@@

\begin{table}[ht]
    \begin{center}
    \begin{tabular}{ | l | l | }
    \hline
    \textbf{Algorithm} & \textbf{Error rate} \\ \hline
    CN2 & 15.20 \\ \hline
    ID3 & 15.68 \\ \hline
    ADTree & 18.49 \\ \hline
    C4.5 & 18.84 \\ \hline
    Bayesian Networks (Simulated Annealing) & 19.15 \\ \hline
    Bayesian Networks (Hill Climbing) & 20.03 \\ \hline
    ILP (Aleph) & 20.45 \\ \hline
    Random Forest & 21.01 \\ \hline
    Bayesian Networks (Genetic Search) & 21.26 \\ \hline
    CART & 21.34 \\ \hline
    \end{tabular}
    \caption{Error rates for the different algorithms used in the classification process}
    \label{tbl:overall_results}
    \end{center}
\end{table}

@@

\caption{Summary of the value distribution on the training data of the dataset}

@@

       Mean & 38.77 \\
       private & 11210 \\
       self\_emp\_not\_inc & 1321 \\
       local\_gov & 1043 \\
       state\_gov & 683 \\
       self\_emp\_inc & 579 \\
       (Other) & 482 \\
       NA's & 963 \\
       hs\_grad & 5283 \\
       some\_college & 3587 \\
       bachelors & 2670 \\
       masters & 934 \\
       assoc\_voc & 679 \\
       11th & 637 \\
       (Other) & 2491 \\
       divorced & 2190 \\
       married\_af\_spouse & 14 \\
       married\_civ\_spouse & 7403 \\
       married\_spouse\_absent & 210 \\
       never\_married & 5434 \\
       separated & 505 \\
       widowed & 525 \\
       prof\_specialty & 2032 \\
       exec\_managerial & 2020 \\
       craft\_repair & 2013 \\
       sales & 1854 \\
       adm\_clerical & 1841 \\
       (Other) & 5555 \\
       NA's & 966 \\
       husband & 6523 \\
       not\_in\_family & 4278 \\
       other\_relative & 525 \\
       own\_child & 2513 \\
       unmarried & 1679 \\
       wife & 763 \\
       amer\_indian\_eskimo & 159 \\
       asian\_pac\_islander & 480 \\
       black & 1561 \\
       other & 135 \\
       white & 13946 \\
       female & 5421 \\
       male & 10860 \\
       Mean & 1082 \\
       Mean & 87.9 \\
       Max. & 3770 \\
       Mean & 40.39 \\
       United States & 14662 \\
       Mexico & 308 \\
       Philippines & 97 \\
       Puerto Rico & 70 \\
       Germany & 69 \\
       (Other) & 801 \\
       NA's & 274 \\
       No & 12435 \\
       Yes & 3846 \\
\caption{Summary of the value distribution on the test data of the dataset}
\label{tab:summarytest}
